---
execute: 
  echo: true
  eval: false
  warning: false
---

# Motif-based promoter similarities between paralogs

Here, we will quantify similarities between promoters of paralogous genes.
Our goal is to better understand how different duplication modes preserve
cis-regulatory landscapes.

We will start by loading the required packages.

```{r here}
#| message: false
#| eval: true

library(here)
library(Biostrings)
library(GenomicRanges)
library(monaLisa)
library(JASPAR2024)
library(TFBSTools)
library(tidyverse)

set.seed(123) # for reproducibility
options(timeout = 1e6) # to download large data files

# Load helper functions
source(here("code", "utils.R"))

# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))

# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

We will also load some required objects created in previous chapters.

```{r}
# Read duplicated gene pairs with age-based group classifications
pairs_age <- readRDS(
    here("products", "result_files", "pairs_by_age_group_expanded.rds")
)
```

```{r}
#| echo: false
#| eval: true

# Load pre-computed plots to display after they are created
f <- list.files(here("products", "plots"), full.names = TRUE)
f <- f[!grepl("one_slide_per", f)]
plots <- lapply(f, readRDS)
names(plots) <- gsub("\\.rds", "", basename(f))
```

## Obtaning promoter sequences

We will first obtain promoter sequences, hereafter defined as the genomic
intervals between -1000 bp and +200 bp.

```{r}
# Define paths to data files
data_list <- list(
    ath = c(
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/fasta/arabidopsis_thaliana/dna/Arabidopsis_thaliana.TAIR10.dna_rm.toplevel.fa.gz",
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/gff3/arabidopsis_thaliana/Arabidopsis_thaliana.TAIR10.61.gff3.gz"
    ),
    gma = c(
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/fasta/glycine_max/dna/Glycine_max.Glycine_max_v2.1.dna_rm.toplevel.fa.gz",
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/gff3/glycine_max/Glycine_max.Glycine_max_v2.1.61.gff3.gz"
    ),
    pap = c(
        "https://orchidstra2.abrc.sinica.edu.tw/orchidstra2/pagenome/padownload/P_aphrodite_genomic_scaffold_v1.0.fa.gz",
        "https://orchidstra2.abrc.sinica.edu.tw/orchidstra2/pagenome/padownload/P_aphrodite_genomic_scaffold_v1.0_gene.gff.gz"
    ),
    zma = c(
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/fasta/zea_mays/dna/Zea_mays.Zm-B73-REFERENCE-NAM-5.0.dna_rm.toplevel.fa.gz",
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/gff3/zea_mays/Zea_mays.Zm-B73-REFERENCE-NAM-5.0.61.gff3.gz"
    ),
    hvu = c(
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/fasta/hordeum_vulgare/dna/Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna_rm.toplevel.fa.gz",
        "https://ftp.ebi.ac.uk/ensemblgenomes/pub/release-61/plants/gff3/hordeum_vulgare/Hordeum_vulgare.MorexV3_pseudomolecules_assembly.61.gff3.gz"
    )
)

# Get promoter sequences based on genome and annotation
prom_seqs <- lapply(seq_along(data_list), function(n) {
    
    sp <- names(data_list)[n]
    
    # Load genome and annotation
    genome <- readDNAStringSet(data_list[[sp]][1])
    names(genome) <- gsub(" .*", "", names(genome))
    
    annot <- rtracklayer::import(data_list[[sp]][2])
    
    # Process annotation
    annot <- annot[annot$type == "gene"]
    sl <- setNames(width(genome), names(genome))
    sl <- sl[seqlevels(annot)]
    seqlengths(annot) <- sl
    
    ## Extract promoter sequences
    prom_ranges <- trim(promoters(annot, 1000, 200))
    prom_seqs <- BSgenome::getSeq(genome, prom_ranges)
    if(sp == "pap") {
        names(prom_seqs) <- prom_ranges$ID
    } else {
        names(prom_seqs) <- prom_ranges$gene_id
    }
    
    ## Keep only duplicated genes
    sel_genes <- unique(c(pairs_age[[sp]]$dup1, pairs_age[[sp]]$dup2))
    prom_seqs <- prom_seqs[sel_genes]
    
    fpath <- tempfile(fileext = ".fa.gz")
    writeXStringSet(prom_seqs, filepath = fpath)
    fprom_seqs <- readDNAStringSet(fpath)
    
    unlink(fpath)
    
    return(fprom_seqs)
})
names(prom_seqs) <- names(data_list)
```

## Annotating promoter sequences with TFBMs

First, we will obtain non-redundant plant TFBMs from 
JASPAR2024 [@rauluseviciute2024jaspar].

```{r}
# Retrieve data from JASPAR2024 ----
js24 <- RSQLite::dbConnect(RSQLite::SQLite(), db(JASPAR2024()))

## PWMs and PFMs ----
unopts <- list(collection = "CORE", tax_group = "plants")
pwms <- getMatrixSet(js24, opts = c(unopts, matrixtype = "PWM"))

## Metadata ----
motif_meta <- Reduce(rbind, lapply(pwms, function(x) {
    df <- tryCatch(
        data.frame(
            id = ID(x),
            name = name(x),
            family = tags(x)$family,
            class = matrixClass(x),
            species = unname(tags(x)$species)
        ),
        error = function(e) return(NULL)
    )
    return(df)
}))

## Pairwise motif comparisons
motif_comp <- read_tsv(
    "https://jaspar.elixir.no/static/clustering/2024/plants/CORE/interactive_trees/pairwise_comparisons.tab", show_col_types = FALSE, skip = 1
) |>
    as.data.frame() |>
    mutate(
        id1 = str_replace_all(`#id1`, ".*_CORE_", ""),
        id1 = str_replace_all(id1, "_n.*", ""),
        id2 = str_replace_all(id2, ".*_CORE_", ""),
        id2 = str_replace_all(id2, "_n.*", "")
    ) |>
    select(id1, id2, cor) 

comp_mat <- motif_comp |>
    igraph::graph_from_data_frame() |>
    igraph::as_adjacency_matrix(attr = "cor") |>
    as.matrix()

# Remove redundancy (similarity >0.9)
tf_fams <- unique(motif_meta$family)
nr_motifs <- Reduce(rbind, lapply(tf_fams, function(x) {
   
    ## Get similarity matrix with selected motifs only
    sel <- motif_meta |> filter(family == x) |> pull(id)
    
    mclusters <- setNames(rep(1, length(sel)), sel)
    if(length(sel) >1) {
        fmat <- comp_mat[sel, sel]
        
        ## Get clusters
        mclusters <- hclust(as.dist(1 - fmat))
        mclusters <- cutree(mclusters, h = 0.1)
    }

    df <- data.frame(
        id = names(mclusters),
        cluster = paste0(x, "_", mclusters),
        family = x
    ) |>
        arrange(cluster)
    
    return(df)
})) |>
    distinct(cluster, .keep_all = TRUE) |>
    pull(id)

fpwms <- pwms[nr_motifs]
```

Now, we will annotate promoters with motifs.

```{r}
# Find motif hits
motif_hits <- lapply(prom_seqs, function(x) {
    return(
        findMotifHits(
            query = fpwms,
            subject = x,
            min.score = "90%",
            method = "matchPWM"
        ) |>
            as.data.frame()
    )
})
```

## Calculating promoter similarity between paralogs

Now that we have a data frame summarizing what TFBMs are found in promoters
of all duplicated genes, we will calculate Sorensen-Dice similarities 
(single and multiset) between paralogs.

```{r}
paralogs <- pairs_age$hvu

# Calculate single and multiset Sorensen-Dice similarities
paralogs_sim <- lapply(names(motif_hits), function(sp) {
    
    paralogs <- pairs_age[[sp]]
    ms_sdice <- lapply(seq_len(nrow(paralogs)), function(n) {
        
        m1 <- motif_hits[[sp]] |> filter(seqnames == paralogs[n, 1]) |> pull(pwmid)
        m2 <- motif_hits[[sp]] |> filter(seqnames == paralogs[n, 2]) |> pull(pwmid)
        
        ## Single Sorensen-Dice similarity
        m1u <- unique(m1)
        m2u <- unique(m2)
        sd_single <- 2 * length(intersect(m1u, m2u)) / (length(m1u) + length(m2u))
        
        ## Multiset Sorensen-Dice similarity
        tab1 <- table(m1)
        tab2 <- table(m2)
        common <- intersect(names(tab1), names(tab2))
        int_count <- sum(pmin(tab1[common], tab2[common]))
        sd_multi <- 2 * int_count / (length(m1) + length(m2))
        
        df_sim <- data.frame(sd_single = sd_single, sd_multi = sd_multi)
        return(df_sim)
    }) |> 
        bind_rows() |>
        mutate(species = sp)
    
    paralogs_final <- cbind(paralogs, ms_sdice)
    
    return(paralogs_final)
}) |>
    bind_rows() |>
    mutate(species_peak = str_c(species, peak_name, sep = "_"))
```

Then, we will visualize a distribution of promoter similarities by
duplication modes.

```{r}
# Compare distros and get CLDs
cld_df <- lapply(
    split(paralogs_sim, paralogs_sim$species_peak),
    cld_kw_dunn, value = "sd_multi"
) |>
    bind_rows(.id = "species_peak") |>
    separate_wider_delim(species_peak, "_", names = c("species", "peak_name")) |>
    mutate(
        type = factor(Group, levels = names(dup_pal)),
        species = str_to_title(species),
        species = factor(species, levels = c("Ath", "Gma", "Pap", "Zma", "Hvu"))
    )
    
# Plot violin + boxplots with CLDs
p_prom_sim <- paralogs_sim |>
    mutate(
        species = str_to_title(species),
        species = factor(species, levels = c("Ath", "Gma", "Pap", "Zma", "Hvu"))
    ) |>
    ggplot(aes(x = sd_multi, y = type)) +
    geom_violin(aes(fill = type)) +
    scale_fill_manual(values = dup_pal) +
    geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
    geom_label(
        data = cld_df,
        aes(x = 0.85, y = type, label = Letter)
    ) +
    ggh4x::facet_nested(
        cols = vars(species, peak_name), space = "free"
    ) +
    scale_x_continuous(
        limits = c(0, 1), breaks = c(0, 0.5, 1), labels = c("0", "0.5", "1")
    ) +
    theme_classic() +
    theme(
        panel.background = element_rect(fill = bg), legend.position = "none"
    ) +
    labs(
        title = "Motif-based promoter similarities between paralogs",
        x = "Similarity", y = "Mode"
    )

p_prom_sim
```

```{r}
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4
#| fig-dpi: 300

plots$promoter_similarity_distros +
    labs(
        title = "Motif-based promoter similarities between paralogs",
        x = "Similarity", y = "Mode"
    )
```

The figure shows that SD, TD, and PD duplicates display higher motif 
similarities, as expected. However, it's also important to note that
promoters of dTRD duplicates are often more similar than promoters of DD
and rTRD duplicates, indicating that gene duplication through DNA 
transpositions can preserve cis-regulatory landscapes to some extent 
(possibly by copying genes and a fraction of the promoter sequences), 
although not at the same level of SD and TD.

## Saving objects {.unnumbered}

Finally, we will save important objects to reuse later.

```{r}
#| eval: false

# Objects
## List of data frames with motif hits in promoters
saveRDS(
    motif_hits, compress = "xz",
    file = here("products", "result_files", "motif_hits_promoters.rds")
)

## Data frame of TFBM-based promoter similarities
saveRDS(
    paralogs_sim, compress = "xz",
    file = here("products", "result_files", "promoter_similarities.rds")
)

# Plots
## Distribution of promoter similarities
saveRDS(
    p_prom_sim, compress = "xz",
    file = here("products", "plots", "promoter_similarity_distros.rds")
)
```

## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
